\paperTitle

### 2014-05-12 doc/icml2014/directed.tex
gotten up to lemma 2

% DONE: need more setup of what the goal is
%The tensor factorization method attacks the heart of the non-convexity
  %in latent variable models.
  In this section, we will develop a consistent parameter estimate for a class of directed graphical models.
  In general directed models, the conditional moments $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i)$
  do not correspond directly to the underlying parameters.
  %Once we recover the conditional moments $\mOpp{v}{j} \eqdef \Pr(x_v
  %| h_j)$ for some $v, j$, we present a systematic approach to learn the
  %conditional probability tables for every clique for a class of directed
  %graphical models.
  Let us see how we can estimate the parameters with an example (Section~\ref{sec:directedExample}).
  Section~\ref{sec:directedGeneral} will describe the algorithm in full generality.
For ease of exposition we make the following simplifications to our presentation:
(i) we describe our algorithm entirely in the context of directed
  graphical models, though it generalizes to factor graphs;
(ii) we derive our algorithm assuming infinite data; in \sectionref{sampleComplexity} we 
  describe error bounds for the finite sample regime; and
(iii) we present the algorithm solely in terms of linear algebraic operations; in
  \sectionref{piecewise}, we show how we can get a more statistically
  efficient estimator using composite likelihoods.

\subsection{Example: Directed grid model}
\label{sec:directedExample}

@@

%Consider the directed grid model shown in \figureref{grid} which
Consider the directed grid model shown in \figureref{approach}.
  % DONE: all the different dependency => too strong
  %captures the local dependency structures possible in
  %a Bayesian network.
The model has eight observed variables $x^a_1, x^b_1 \cdots, x^a_4, x^b_4$ and four
  hidden variables $h_1, \cdots, h_4$.
The parameters of this model are the conditional probability tables
$\pi \eqdef \Pr(h_1) \in \Re^k, T \eqdef \Pr(h_2 | h_1) = \Pr(h_3 | h_1) \in \Re^{k \times k},
V \eqdef \Pr(h_4 | h_2, h_3) \in \Re^{k \times k \times k}$ and $O \eqdef \Pr(x^a_i | h_i)
=  \Pr(x^b_i | h_i) \in \Re^{d \times k}$. 
Assume $O$ and $T$ have full column rank.

\paragraph{Estimating $O$}
Note that the observed variables $x^a_1, x^b_1, x^a_2$ are
  conditionally independent given $h_1$; we can use
  $\TensorFactorize$ (\sectionref{setup}) to recover $O$.

\paragraph{Estimating $\pi$}
The moments of $x^a_1$, $\mO_1 \eqdef \Pr(x^a_1)$ are directly related to
  $\pi$ by $\mO_1 = O \pi$. 
By \assumptionref{full-rank}, $O$ has full column rank and thus can be
  inverted to recover $\pi$: $\pi = O\pinv  \mO_1$.

\paragraph{Estimating $T$}
Similarly, we can write down the moments of $x^a_1, x^a_2$, $\mO_{12}
  \eqdef \Pr(x^a_1, x^a_2)$, in terms of the hidden marginals $\mH_{12}
  \eqdef \Pr(h_1, h_2)$ and solve for $\mH_{12}$:
\begin{align*}
\mO_{12} = O \mH_{12} O^\top \quad\Rightarrow\quad
  \mH_{12} = O\pinv  \mO_{12} O\pinvt .
\end{align*}
$T$ can be recovered from the $\mH_{12}$ by suitable renormalization.

@@

Note that in each of the above cases,
  for any clique $\sC$, each hidden variable $h_i \in \sC$ has its own
  view which is conditionally independent given the other hidden variables in
  the clique.
%Fundamentally, algorithm takes advantage of having a view for every
  %hidden variable in a clique to help identify its hidden state.
We capture this general property as follows:
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
A hidden clique $\sC \in \sG$ is said to possess the \textbf{exclusive views property} if for
  every hidden variable $h_i \in \sC$, there is some observed variable,
  $x_{v}$ which is conditionally independent of the rest of the clique
  given $h_i$, and whose conditional moment $\mOpp{v}{i} \eqdef
  \Pr(x_{v} \mid h_i)$ can be estimated consistently.
We call $x_v$ the exclusive view of $h_i$ in $\sC$.

This property also extends to any connected sub-graph $\sG' \in \sG$.
\end{property}
For the rest of this section, we will only consider sub-graphs which are
  cliques. 
In \sectionref{undirected}, we will show how the generalization
  to sub-graphs allows us to efficiently estimate the pseudo-likelihood.

@@

We now show this condition is sufficient to learn the marginal
  distribution of the clique.
Consider any hidden clique $\sC = \{i_1, \cdots, i_m\}$ for which this property holds. Let
  $x_{v_j}$ be any exclusive view for $h_{i_j}$ in $\sC$ and $\sV
  = \{v_1, \cdots, v_m\}$ be a set of exclusive views for the clique $\sC$.
We can write down the moments $\mO_\sV$ as follows:
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  %&= \sum_{\vh \in \sH}
      %\Pr(\Sh{\sC}) \Pr(\Sx{\sV} \given \Sh{\sC}) \\
      &= \sum_{\vh \in \sH} \Pr(\Sh{\sC}) 
          \Pr(x_{v_1} | h_{i_1}) \cdots \Pr(x_{v_m} | h_{i_m}) \\
    &= Z_{\sC}(\mOpp{v_1}{i_1},\cdots,\mOpp{v_m}{i_m}).
\end{align*}

If each $\mOpp{v_j}{i_j}$ has full column rank, then we can recover the
hidden marginals $Z_\sC$ by inverting:
\begin{align*}
  Z_{\sC} &= \mO_\sV(\mOpp{v_1}{i_1}\pinv,\cdots,\mOpp{v_m}{i_m}\pinv).
\end{align*}
Finally, the conditional probability tables for $\sC$ can easily be obtained from
  $Z_\sC$ via renormalization.
\algorithmref{learnclique} summarizes the procedure, \LearnClique.

\begin{algorithm}
  \caption{\LearnClique~(pseudoinverse)}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE Hidden clique $\sC$ with exclusive views (\propertyref{exclusive-views}).
    \ENSURE Marginal distribution of the clique $Z_\sC$.
      \STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \cdots, x_{v_m}\}$.
      \STATE Return $\mH_\sC \gets \mO_{\sV}( \mOpp{v_1}{i_i}\pinv, \cdots, {\mOpp{v_m}{i_m}}\pinv )$.
  \end{algorithmic}
\end{algorithm}
Of course, for $Z_{\sC}$ to be identifiable, we need various rank conditions,
which will be established in \sectionref{sampleComplexity}.

@@

% DONE: transition the content more gradually (checkpoint: what have we done so far, what next)
% DONE: refer to properties by name in addition to by number
We have shown how to estimate the parameters of any clique that possesses the exclusive
  views property (\propertyref{exclusive-views}). But in a general graph, which cliques have this property?
In the example above, we were able to guarantee \propertyref{exclusive-views}
  by identifying a hidden variable, $h_1$ along with three conditionally
  independent observed variables, $x^a_1, x^b_1, x^a_2$ using the tensor
  factorization method.
  %followed by using \TensorFactorize to learn the conditional moments.
We define such a hidden variable to be a bottleneck:
\begin{definition}[Bottleneck]
  A hidden variable $h$ is said to be a \textbf{bottleneck} if there
  exists at least three observed variables, or views, $x_1, x_2, x_3$
  that are conditionally independent given $h$. 
  Let $\sV_{h}$ denote a set of views for $h$ (not necessarily unique).
\end{definition}

% DONE: don't talk about parameter recovery yet, just establish exclusive views
%Finally, the claim we make is that we can recover parameters for any
  %graphical model where every hidden variable is a bottleneck:
We will prove that the following simple property guarantees that all cliques
have the exclusive views property:
\begin{property}[Uniformly bottlenecked]
  \label{prop:bottleneck}
  Every hidden variable is a bottleneck.
\end{property}

@@

Before we establish this relationship,
we describe a reduction of any graphical model to
  a canonical form to make our arguments simpler.
  % DONE: don't say this yet since reader can't appreciate it.
  %that will allow us to handle the case where an observed
  %variable has multiple parents.

@@

  \providecommand{\hp}{\ensuremath{h_\text{\rm new}}}

  Let $x_v$ be an observed variable with parents $\Pa(x_v)$ and children $\text{Ch}(x_v)$.
  Consider the following transformation.
  Replace $x_v$ with a new hidden variable \hp\ with the same
  parents $\Pa(x_v)$ and children $\text{Ch}(x_v) \union \{x_{v_1}, x_{v_2}, x_{v_3}\}$,
  where $x_{v_1},x_{v_2},x_{v_3}$ are three copies of $x$
  (\figureref{reduction}). 
  The domain of $\hp$ is the same as $x_v$\footnote{
      Though we have assumed that all hidden variables share the
      same domain, $[k]$, our work generalizes to hidden domains of
      different sizes, provided \assumptionref{full-rank} is satisfied.
      },
    and $\mOpp{v}{\textrm{new}} = I$.
  Then, there is a one-to-one correspondence between every value of
  $\hp$ and $x_v$. Consequently, for any clique $\sC \contains \hp$, the
  parameters in the original graphical model can be obtained by
  substituting $\hp$ with $x_v$.

@@

\begin{lemma}[Bottlenecks guarantee exclusive views]
  \label{lem:bottleneck-views}  
Let $\sG' = \{ h_{i_1}, \cdots, h_{i_m} \} \subseteq \sG$ be
  a connected sub-graph of hidden variables. If every hidden variable
  $h_i \in \sG'$ is a bottleneck, then they have an exclusive view.
\end{lemma}
\begin{proof}
  The independence conditions required by the bottleneck condition
  effectively guarantee that atleast one of the three views for
  a bottleneck be exclusive. 
  \appendixref{exclusive-views} provides a complete proof.
\end{proof}

@@

      \FOR{each hidden variable $h_i \in H$} 
        \STATE Apply $\TensorFactorize$ to learn conditional moments
        $\mOpp{v}{i}$ for every $\sV_{h_i}$.

%    \COMMENT{Recover observation potentials $O$ using bottlenecks}
      \ENDFOR
%      \COMMENT{\textbf{Step 2:} Recover clique potentials from the piecewise likelihood.}
\FOR{every clique $\sC = \{h_{i_1}, \cdots, h_{i_m}\} \in \sG$} 
\STATE Apply $\LearnClique$ to learn the marginals $\mH_\sC$.
\ENDFOR
  \end{algorithmic}
\end{algorithm}

>>>>

% DONE: need more setup of what the goal is
%The tensor factorization method attacks the heart of the non-convexity
  %in latent variable models.
  In this section, we will develop a consistent parameter estimate for a class of directed graphical models.
  In general directed models, the conditional moments $\mOpp{v}{i} \eqdef \Pr(x_v \mid h_i)$
which we were able to recover from {\TensorFactorize}, are not the underlying parameters.
  %Once we recover the conditional moments $\mOpp{v}{j} \eqdef \Pr(x_v
  %| h_j)$ for some $v, j$, we present a systematic approach to learn the
  %conditional probability tables for every clique for a class of directed
  %graphical models.
  Let us first consider an example (Section~\ref{sec:directedExample});
  Section~\ref{sec:directedGeneral} will describe the algorithm in full generality.
For ease of exposition we make the following simplifications to our presentation:
(i) we describe our algorithm entirely in the context of directed
  graphical models, though it generalizes to undirected models;
(ii) we derive our algorithm assuming infinite data; in \sectionref{sampleComplexity} we 
  describe error bounds for the finite sample regime; and
(iii) we present the algorithm solely in terms of linear algebraic operations;
  \sectionref{piecewise} provides a statistically
  more efficient estimator using composite likelihood.

\subsection{Example: directed grid model}
\label{sec:directedExample}

@@

%Consider the directed grid model shown in \figureref{grid} which
Consider the directed grid model shown in \figureref{approach}.
  % DONE: all the different dependency => too strong
  %captures the local dependency structures possible in
  %a Bayesian network.
The model has eight observed variables $x^a_1, x^b_1 \cdots, x^a_4, x^b_4$ and four
  hidden variables $h_1, \ldots, h_4$.
The parameters of this model are the conditional probability tables
$\pi \eqdef \Pr(h_1) \in \Re^k, T \eqdef \Pr(h_2 | h_1) = \Pr(h_3 | h_1) \in \Re^{k \times k},
V \eqdef \Pr(h_4 | h_2, h_3) \in \Re^{k \times k \times k}$ and $O \eqdef \Pr(x^a_i | h_i)
=  \Pr(x^b_i | h_i) \in \Re^{d \times k}$. 
%Assume $O$ and $T$ have full column rank. % PL: if we put this in, we have to say something about V, which is too complex.

\paragraph{Estimating $O$}
Note that the observed variables $x^a_1, x^b_1, x^a_2$ are
  conditionally independent given $h_1$; we can therefore use
  $\TensorFactorize$ from \sectionref{setup} to recover $O$.

\paragraph{Estimating $\pi$}
The moments of $x^a_1$, $\mO_1 \eqdef \Pr(x^a_1)$ are directly related to
  $\pi$ by a linear system: $\mO_1 = O \pi$. 
% PL: weird to refer to refer to this assumption from previous section
% where it's talking about another model.
%By \assumptionref{full-rank}, $O$ has full column rank and thus can be
  %inverted to recover $\pi$: $\pi = O\pinv  \mO_1$.
If $O$ has full column rank, we can recover $\pi$ by taking the pseudoinverse: $\pi = O\pinv  \mO_1$.

\paragraph{Estimating $T$}
Similarly, we can write down the moments of $x^a_1, x^a_2$, $\mO_{12}
  \eqdef \Pr(x^a_1, x^a_2)$, in terms of the hidden marginals $\mH_{12}
  \eqdef \Pr(h_1, h_2)$ and solve for $\mH_{12}$:
\begin{align*}
\mO_{12} = O \mH_{12} O^\top \quad\Rightarrow\quad
  \mH_{12} = O\pinv  \mO_{12} O\pinvt .
\end{align*}
$T$ can be recovered from the $\mH_{12}$ by renormalizing the columns.

@@

%Let us now generalize the intuitions of the directed grid to general graphs.
The intuitions of the directed grid generalize readily to general graphs.
% PL: this prose is redundant with property
%Note that in each of the above cases,
  %for any clique $\sC$, each hidden variable $h_i \in \sC$ has its own
  %view which is conditionally independent given the other hidden variables in
  %the clique.
%Fundamentally, algorithm takes advantage of having a view for every
  %hidden variable in a clique to help identify its hidden state.
The key property required for our algorithm to work is as follows:
%We capture this general property as follows:
\begin{property}(Exclusive views)
  \label{prop:exclusive-views}
A hidden clique $\sC \in \sG$ is said to possess the \emph{exclusive views property} if for
  every hidden variable $h_i \in \sC$, (i)
  there exists some observed variable,
  $x_{v}$ which is conditionally independent of the rest of the clique
  given $h_i$, and (ii) the conditional moments $\mOpp{v}{i} \eqdef
  \Pr(x_{v} \mid h_i)$ can be estimated consistently.
We call $x_v$ an exclusive view of $h_i$ in $\sC$.
If every clique has exclusive views, then we say that $\sG$ does as well.
\end{property}
% PL: to save space
%This property also extends to any connected subgraph $\sG' \in \sG$.
%For the rest of this section, we will only consider subgraphs which are
  %cliques. 
%In \sectionref{undirected}, we will show how the generalization
  %to subgraphs allows us to efficiently estimate the pseudo-likelihood.

@@

We now show this condition is sufficient to learn the marginal
  distribution of the clique.
Consider any hidden clique $\sC = \{i_1, \ldots, i_m\}$ with the exclusive views property. Let
  $x_{v_j}$ be any exclusive view for $h_{i_j}$ in $\sC$ and define $\sV
  = \{v_1, \ldots, v_m\}$. % be a set of exclusive views for the clique $\sC$.
We can write down the moments $\mO_\sV$ as follows:
\begin{align*}
  \mO_\sV 
  &\eqdef \Pr(\Sx{\sV}) \\
  %&= \sum_{\vh \in \sH}
      %\Pr(\Sh{\sC}) \Pr(\Sx{\sV} \given \Sh{\sC}) \\
      &= \sum_{\vh_\sC} \Pr(\Sh{\sC}) 
          \Pr(x_{v_1} | h_{i_1}) \cdots \Pr(x_{v_m} | h_{i_m}) \\
    &= Z_{\sC}(\mOpp{v_1}{i_1},\cdots,\mOpp{v_m}{i_m}).
\end{align*}

If each $\mOpp{v_j}{i_j}$ has full column rank, then we can recover the
hidden marginals $Z_\sC$ by inverting:
\begin{align*}
  Z_{\sC} &= \mO_\sV(\mOpp{v_1}{i_1}\pinv,\cdots,\mOpp{v_m}{i_m}\pinv).
\end{align*}
\algorithmref{learnclique} summarizes the procedure, \LearnClique.
Given $Z_\sC$, the conditional probability tables for $\sC$ can easily be
obtained via renormalization (see \sectionref{sampleComplexity}).

\begin{algorithm}
  \caption{\LearnClique~(pseudoinverse)}
  \label{algo:learnclique}
  \begin{algorithmic}
    \REQUIRE Hidden clique $\sC$ with exclusive views (\propertyref{exclusive-views}).
    \ENSURE Marginal distribution of the clique $Z_\sC$.
      \STATE Identify exclusive views $\Sx{\sV} = \{x_{v_1}, \dots, x_{v_m}\}$.
      \STATE Return $\mH_\sC \gets \mO_{\sV}( \mOpp{v_1}{i_i}\pinv, \dots, {\mOpp{v_m}{i_m}}\pinv )$.
  \end{algorithmic}
\end{algorithm}
% PL: sample complexity, not rank
%Of course, for $Z_{\sC}$ to be identifiable, we need various rank conditions,
%which will be established in \sectionref{sampleComplexity}.

@@

% DONE: transition the content more gradually (checkpoint: what have we done so far, what next)
% DONE: refer to properties by name in addition to by number
We have shown how to estimate the parameters of any clique that possesses the exclusive
  views property (\propertyref{exclusive-views}).
  But in a general graph, which cliques have this property?
  To provide intuition, we will provide more interpretable sufficient conditions
  and examples.

%In the example above, we were able to guarantee \propertyref{exclusive-views}
  %by identifying a hidden variable, $h_1$ along with three conditionally
  %independent observed variables, $x^a_1, x^b_1, x^a_2$ using the tensor
  %factorization method.
  %followed by using \TensorFactorize to learn the conditional moments.
First, let us revisit the definition of a bottleneck:
\begin{definition}[Bottleneck]
  A hidden variable $h_i$ is said to be a \textbf{bottleneck} if there
  exists at least three observed variables (views), $x_{v_1}, x_{v_2}, x_{v_3}$
  that are conditionally independent given $h_i$,
  and $\mOpp{v_1}{i},\mOpp{v_2}{i},\mOpp{v_3}{i}$ have full column rank $k$.
  % PL: need to add rank constraint; otherwise, doesn't imply exclusive views
  Let $\sV_{h_i}$ denote any set of views for $h_i$.
\end{definition}

% DONE: don't talk about parameter recovery yet, just establish exclusive views
%Finally, the claim we make is that we can recover parameters for any
  %graphical model where every hidden variable is a bottleneck:
Now define the following property:
\begin{property}[Uniformly bottlenecked]
  \label{prop:bottleneck}
  A graphical model $\sG$ is \emph{uniformly bottlenecked} if
  every hidden variable in $\sG$ is a bottleneck.
\end{property}
Note that two hidden variables are allowed to have overlapping views.
%If every hidden variable is a bottleneck, then we can estimate $\mOpp{v}{i}$
It is clear that having uniform bottlenecks implies part (ii)
of the exclusive views property (\propertyref{exclusive-views}) via \TensorFactorize,
but we will show that part (i) follows as well.

%We will prove that the following simple property guarantees that all cliques
%have the exclusive views property:

@@

%Before we establish this relationship,
First, we describe a reduction of any graphical model to
  a canonical form to make our arguments simpler.
  % DONE: don't say this yet since reader can't appreciate it.
  %that will allow us to handle the case where an observed
  %variable has multiple parents.

@@

  \providecommand{\hp}{\ensuremath{h_\text{\rm new}}}
  Let $x_v$ be an observed variable with parents $\Pa(x_v)$ and children $\text{Ch}(x_v)$.
  Consider the following transformation.
  Replace $x_v$ with a new hidden variable \hp\ with the same
  parents $\Pa(x_v)$ and children $\text{Ch}(x_v) \union \{x_{v_1}, x_{v_2}, x_{v_3}\}$,
  where $x_{v_1},x_{v_2},x_{v_3}$ are three copies of $x_v$
  (\figureref{reduction}).
  Define $\Pr(\hp \mid \Pa(x_v)) = \Pr(x_v \mid \Pa(x_v))$ and
  $\Pr(x_{v_j} \mid \hp) = I$.
  %$\Pr(x_{v_j} \mid \hp) = \left[ \begin{array}{c} I_{k \times k} \\ 0_{(d-k) \times k} \end{array} \right]$.
  %The domain of $\hp$ is the same as $x_{v_1},x_{v_2},x_{v_3}$.
  % PL: cut to save space
  %\footnote{
      %Though we have assumed that all hidden variables share the
      %same domain, $[k]$, our work generalizes to hidden domains of
      %different sizes, provided \assumptionref{full-rank} is satisfied.
      %},
    %and $\mOpp{v}{\textrm{new}} = I$.
  Then, there is a one-to-one correspondence between every value of
  $\hp$ and $x_v$. Consequently, for any clique $\sC \contains \hp$, the
  parameters in the original graphical model can be obtained by
  substituting $\hp$ with $x_v$.

@@

Now we establish the link between bottlenecks and exclusive views
(see \appendixref{exclusive-views} for the complete proof):
\begin{lemma}[Uniformly bottlenecked implies exclusive views]
  \label{lem:bottleneck-views}  
% PL: simplify to cliques (this is also bad notation, since \sG is a set of cliques, not nodes)
%Let $\sG' = \{ h_{i_1}, \dots, h_{i_m} \} \subseteq \sG$ be
  %a connected subgraph of hidden variables.
  %If every hidden variable
  %$h_i \in \sG'$ is a bottleneck, then $\sG'$ has the exclusive views property.
If a graphical model $\sG$ is uniformly bottlenecked,
then $\sG$ has the exclusive views property.
\end{lemma}
% PL: save space, doesn't add much
%\begin{proof}
%  The independence conditions required by the bottleneck condition
%  effectively guarantee that at least one of the three views for
%  a bottleneck be exclusive. 
%  \appendixref{exclusive-views} provides a complete proof.
%\end{proof}

@@

      \FOR{each hidden variable $h_i \in H$} 
        \STATE Apply $\TensorFactorize$ to learn conditional moments
        $\mOpp{v}{i}$ for every $v \in \sV_{h_i}$.

%    \COMMENT{Recover observation potentials $O$ using bottlenecks}
      \ENDFOR
%      \COMMENT{\textbf{Step 2:} Recover clique potentials from the piecewise likelihood.}
\FOR{every clique $\sC = \{h_{i_1}, \ldots, h_{i_m}\} \in \sG$} 
\STATE Apply $\LearnClique$ to learn the marginals $\mH_\sC$.
\ENDFOR
  \end{algorithmic}
\end{algorithm}

### 2014-05-12 doc/icml2014/exclusive-views.tex
Added figure to assumption proof
gotten up to lemma 2

\subsection{\lemmaref{lem:bottleneck-views}}
\label{app:exclusive-views}

In \sectionref{directed}, we stated that bottlenecks guaranteed
  exclusive views. We provide the proof below.
\begin{lemma*}[Bottlenecks guarantee exclusive views]
  %\label{lem:bottleneck-views}  
Let $\sG' = \{ h_{i_1}, \cdots, h_{i_m} \} \subseteq \sG$ be
  a connected sub-graph of hidden variables. If every hidden variable
  $h_i \in \sG'$ is a bottleneck, then they have an exclusive view.
\end{lemma*}
\begin{proof}
  By \lemmaref{reduction}, we can assume that every observed
  variable is a leaf in the graph, causing any v-structures in the graph
  to be active. This allows us to reason about independence simply by
  checking whether or not two nodes are connected or not. Note that this
  proof applies for undirected graphs as well.

  Let $h$ be a hidden variable in $\sG'$, let $\sV_{h}$ be the set of
  views for it and let ${\sG'}^- \eqdef \sG' \setminus \{h\}$.
  By definition, the views of $h$ are conditionally independent, viz.
    every connecting path between two views $x_1, x_2 \in \sV_h$ must pass through
    $h$.
  A view $x \in \sV_{h}$ is exclusive for $h$ iff $x \not\in \sV_{h'} ~ \forall h'
  \in {\sG'}^-$.

  Suppose $h$ had no unique view, then $\sV_{h} \subseteq \Union_{h' \in {\sG'}^-} \sV_{h'}$. 
  However, this would imply that for every $x_1, x_2 \in \sV_{h}$, there
  exists some $h_1, h_2 \in {\sG'}^-$ (not necessarily distinct) such that
    $x_1$ is connected to $h_1$ and $x_2$ is connected to $h_2$.
  By participation in the sub-graph, either $h_1 = h_2$ or $h_1$ is
    connected to $h_2$, implying that $x_1$ and $x_2$ are connected
    via a path that does not pass through $h$, contradicting the
    statement that $x_1, x_2 \in \sV_h$.

@@

  Algorithmically, finding the unique view for $h$ can be done by
  subtracting the views of all other variables in the sub-graph from
  $\sV_h$,i.e. $x_{v_h} \in \sV_h \setminus \Union_{h' \in \sG'} \sV_{h'}$
\end{proof}

>>>>

\subsection{\lemmaref{bottleneck-views}}
\label{app:exclusive-views}

In \sectionref{directed}, we stated that bottlenecks guaranteed
  exclusive views. We provide the proof below.
  We generalize the lemma so that it applies
  on any connected subgraph, not just cliques.
  Also, the statement is made in terms of subgraphs,
  not the entire graph.

\begin{lemma*}[Bottlenecks guarantee exclusive views]
  %\label{lem:bottleneck-views}  
Let $\sG' = \{ h_{i_1}, \ldots, h_{i_m} \} \subseteq \sG$ be
  a connected subgraph of hidden variables. If every hidden variable
  $h_i \in \sG'$ is a bottleneck, then $\sG'$ has the exclusive view property.
\end{lemma*}

\begin{proof}
  By \lemmaref{reduction}, we can assume that every observed
  variable is a leaf in the graph, causing any v-structures in the graph
  to be active. This allows us to reason about independence simply by
  checking whether or not two nodes are connected or not.
  Note that this proof applies for undirected graphs as well.

  Let $h$ be a hidden variable in $\sG'$, let $\sV_{h}$ be the set of
  views for it and let ${\sG'}^- \eqdef \sG' \setminus \{h\}$.
  By definition, the views of $h$ are conditionally independent, viz.
    every connecting path between two views $x_1, x_2 \in \sV_h$ must pass through
    $h$.
  A view $x \in \sV_{h}$ is exclusive for $h$ iff $x \not\in \sV_{h'}$ for all $h'
  \in {\sG'}^-$.

  Suppose $h$ had no unique view, then $\sV_{h} \subseteq \Union_{h' \in {\sG'}^-} \sV_{h'}$. 
  However, this would imply that for every $x_1, x_2 \in \sV_{h}$, there
  exists some $h_1, h_2 \in {\sG'}^-$ (not necessarily distinct) such that
    $x_1$ is connected to $h_1$ and $x_2$ is connected to $h_2$.
  By participation in the subgraph, either $h_1 = h_2$ or $h_1$ is
    connected to $h_2$, implying that $x_1$ and $x_2$ are connected
    via a path that does not pass through $h$, contradicting the
    statement that $x_1, x_2 \in \sV_h$.

@@

  Algorithmically, finding the unique view for $h$ can be done by
  subtracting the views of all other variables in the subgraph from
  $\sV_h$,i.e. $x_{v_h} \in \sV_h \setminus \Union_{h' \in \sG'} \sV_{h'}$
\end{proof}

### 2014-05-12 doc/icml2014/intro.tex
Various minor corrections to camera-ready
gotten up to lemma 2

% 1. Latent variable models are good.
Latent-variable graphical models provide compact representations of data
and have been employed across many fields % \todo{find a much better/broader set of citations}
\cite{ghahramani1999variational,jaakkola1999variational,blei03lda,quattoni04crf,beaumont2004bayesian,haghighi06prototype}.
% 2. Learning them is hard.
However, learning these models remains a difficult problem due to the
non-convexity of the negative log-likelihood.  Local methods such as Expectation Maximization (EM)
are the norm, but are susceptible to local optima.

% 3. People have approached unsupervised learning with the MoM magic sauce, but the sauce is limited.
Recently, unsupervised learning techniques based on the spectral method of moments
have offered a refreshing perspective on
this learning problem
\citep{mossel2005learning,
hsu09spectral,
bailly2010spectral,
song2011spectral,
anandkumar11tree,
anandkumar12lda,
anandkumar12moments,
hsu12identifiability,
balle12automata}.
%\todo{make sure we are citing the right people}
These methods exploit the linear algebraic properties of the model to
factorize the moments into parameters, providing strong theoretical guarantees.
However, they apply to a limited set of models, and are thus
not as broadly applicable as EM.

\begin{figure}[t]
  \label{fig:approach}
  \centering
  \subimport{figures/}{approach.tikz}
  \caption{
 % \todo{need three steps: tensor factorization, optimize composite likelihood, recover model parameters;}
  Overview of our approach: we first use tensor factorization to learn the conditional moments for each hidden variable.
  Then we optimize a composite likelihood to recover the hidden clique marginals,
  given which we can extract the model parameters.
  }
\end{figure}

@@

% 4. State what we do: exploit moment constraints to make the problem easier.
In this paper,
we show that a much broader class of discrete directed and undirected graphical models can be consistently estimated,
specifically those in which each hidden variable has three conditionally
independent observed variables (``views'')---see \propertyref{bottleneck}.
The key idea is to leverage the spectral method of moments,
not to directly provide a consistent parameter estimate as in previous work,
but as partial constraints on a likelihood-based objective.
%For a broad class of directed and undirected graphical models

@@

% 5. We get moments from third-order tensors from bottlenecks and factorize them into marginals,
% then get parameters.
% DONE: try to talk about the approach in terms of three steps
Our approach is illustrated in \figureref{approach}. 
First, we identify the views for each hidden variable $h_i$ (for example,
$x_1^a$, $x_1^b$ and $x_3^a$ are conditionally independent given $h_1$) and use
the tensor factorization algorithm of
\citet{anandkumar13tensor} to identify the \emph{conditional
moments} $\Pr[x_i^a \mid h_i]$ and $\Pr[x_i^b \mid h_i]$ for each $i$.
Second, we optimize a \emph{composite marginal likelihood} to recover the marginals over
each hidden clique (e.g., $\mathbb P(h_2, h_3, h_4)$).
Normally, such a marginal likelihood objective would be non-convex,
but given the conditional moments, we obtain a convex objective,
which can be globally optimized using EM.
Finally, the hidden marginals yield the model parameters
via simple post-processing in the directed case and convex optimization in the undirected case.
%Given these factors, we show that there exists a set of observed
  %variables for every clique such that the likelihood is convex in the
  %parameters of that clique. 
%This guarantees that we can recover the true marginal distributions
  %using EM.
  %finally, with some post-processing, we are able to
  %consistently recover the true parameters of the model.
In \sectionref{directed}, we detail our algorithm and
show that our algorithm yields consistent parameter estimates
under non-degenerate rank assumptions on the conditional moments.
%describe the
  %settings and technical assumptions in which this approach is guaranteed
  %to work.
% PL: hard to talk about $O$ and $T$ at this stage.
%As a particular example, we are able to recover parameters for the
  %model depicted in \figureref{approach} if $O$ and $T$ are full-rank.

>>>>

% 1. Latent variable models are good.
Latent-variable graphical models provide compact representations of data
and have been employed across many fields % \todo{find a much better/broader set of citations}
\cite{ghahramani1999variational,jaakkola1999variational,blei03lda,quattoni04crf,beaumont2004bayesian,haghighi06prototype}.
% 2. Learning them is hard.
However, learning these models remains a difficult problem due to the
non-convexity of the negative log-likelihood.  Local methods such as expectation maximization (EM)
are the norm, but are susceptible to local optima.

% 3. People have approached unsupervised learning with the MoM magic sauce, but the sauce is limited.
Recently, unsupervised learning techniques based on the spectral method of moments
have offered a refreshing perspective on
this learning problem
\citep{mossel2005learning,
hsu09spectral,
bailly2010spectral,
song2011spectral,
anandkumar11tree,
anandkumar12lda,
anandkumar12moments,
hsu12identifiability,
balle12automata}.
%\todo{make sure we are citing the right people}
These methods exploit the linear algebraic properties of the model to
factorize moments of the observed data distribution into parameters,
providing strong theoretical guarantees.
However, they apply to a limited set of models, and are thus
not as broadly applicable as EM.

\begin{figure}[t]
  \label{fig:approach}
  \centering
  \subimport{figures/}{approach.tikz}
  \caption{
 % \todo{need three steps: tensor factorization, optimize composite likelihood, recover model parameters;}
  Overview of our approach:
  we first use tensor factorization to learn the \emph{conditional moments}
  for each hidden variable.
  Second, we optimize a composite marginal likelihood to recover the \emph{hidden clique marginals},
  from which we solve for the \emph{model parameters}.
  }
\end{figure}

@@

% 4. State what we do: exploit moment constraints to make the problem easier.
In this paper,
we show that a much broader class of discrete directed and undirected graphical models can be consistently estimated:
specifically those in which \emph{each} hidden variable has three conditionally
independent observed variables (``views'')---see \propertyref{bottleneck}.
Our key idea is to leverage the method of moments,
not to directly provide a consistent parameter estimate as in previous work,
but as constraints on a likelihood-based objective.
%For a broad class of directed and undirected graphical models

@@

% 5. We get moments from third-order tensors from bottlenecks and factorize them into marginals,
% then get parameters.
% DONE: try to talk about the approach in terms of three steps
The essence of our approach is illustrated in \figureref{approach}. 
First, we identify three views for each hidden variable $h_i$ (for example,
$x_1^a$, $x_1^b$ and $x_3^a$ are conditionally independent given $h_1$) and use
the tensor factorization algorithm of
\citet{anandkumar13tensor} to estimate the \emph{conditional
moments} $\Pr(x_i^a \mid h_i)$ and $\Pr(x_i^b \mid h_i)$ for each $i$ (\sectionref{directed}).
Second, we optimize a \emph{composite marginal likelihood} to recover the marginals over
each hidden clique (e.g., $\mathbb P(h_2, h_3, h_4)$).
Normally, such a marginal likelihood objective would be non-convex,
but given the conditional moments, we obtain a convex objective,
which can be globally optimized using EM (\sectionref{piecewise}).
So far, our method has relied only on the conditional independence
structure of the model and applies generically to both directed
and undirected models.
The final step of turning hidden marginals into model parameters
requires some specialization.
In the directed case, this is simple post-processing;
in the undirected case, we need to solve another convex optimization problem (\sectionref{undirected}).
%Given these factors, we show that there exists a set of observed
  %variables for every clique such that the likelihood is convex in the
  %parameters of that clique. 
%This guarantees that we can recover the true marginal distributions
  %using EM.
  %finally, with some post-processing, we are able to
  %consistently recover the true parameters of the model.
%In \sectionref{directed}, we describe our algorithm in more detail and
%show that our algorithm yields consistent parameter estimates
%under non-degenerate rank assumptions.
%describe the
  %settings and technical assumptions in which this approach is guaranteed
  %to work.
% PL: hard to talk about $O$ and $T$ at this stage.
%As a particular example, we are able to recover parameters for the
  %model depicted in \figureref{approach} if $O$ and $T$ are full-rank.
